Individual Variation in Animal Movement and Effects on Seed Dispersal Distance

### 2021-01-08 paper/paper.Rmd
Moved text over to a scrap file

# Methods
## Model description

To study the effects of individual variation in animal movement over final dispersal distances of foraged seeds, we developed a spatially-explicit individual-based model in a homogenous landscape consisting of a single source tree. We incorporated animal movement focusing on two characteristics, a movement distance (MD) sampled from a probability density distribution, and a movement angle (MA). In addition to this, we included gut retention time (GRT), the time that ingested seeds stay within the frugivore until they are dropped. The animal was allowed to move freely within the landscape, and GRT determined when a seed would get dispersed. Once a seed was dropped, its location was recorded and seed dispersal distance was estimated as the distance from the origin to the seed's location. The average seed dispersal distance for each simulation run was calculated and used to estimate seed dispersion, calculated as the mean distance of each seed to the average location of all seeds in the run [@jones_closing_2017]. For each simulation run, the simulation started at the source tree, an animal received a specific number of seeds and each seed was assigned a specific GRT sampled from a gamma distribution [@morales_effects_2006]. The simulation was based on one minute time steps, where at each time step a movement distance (MD) and movement angle (MA) would determine the path for the animal to follow. For each seed, the location where it would get dropped was based on the animal's location at the time it reached the GRT for each of the seeds. The total simulation time for each run was determined by the largest GRT sampled for that specific simulation run. Each simulation run focused on one individual animal dispersing seeds from one focus tree. Once all seeds in that run were dropped by the animal, the simulation run ended.

<!--- I need to include in that first paragraph how we are comparing between models with and without individual variation--->

@@

In order to analyze the effect of differences in animal movement over final dispersal distances of foraged seeds we developed a spatially-explicit individual-based model in a homogeneous landscape that incorporated animal movement distances and seed gut retention time (GRT). Our landscape consisted of a single source tree at the origin,  our simulation time steps were based on one minute intervals, and each simulation started at the source tree (0,0), where each individual bird was given five seeds and was allowed to move freely in the landscape until the simulation ended once all seeds were dropped, Figure1. Each seed's gut retention time was determined at the start of each simulation run, being randomly sampled from a Gamma distribution (shape = 4, scale = 5, shift = 8) [@morales_effects_2006] so that the average gut retention time was 28 minutes[@holbrook_dispersal_2009; @holbrook_using_2007]. For each simulation run, the total simulation time was determined by the largest gut retention time sampled for the five seeds, therefore, once the last seed was dropped, the simulation ended. We calculated the Euclidean distance from the source tree to each of the dropped seeds to determine their seed dispersal distance. We simulated animal movement using a memoryless process, a random walk with a uniformly distributed angular direction and a random step size chosen from a probability distribution. The probability distribution chosen to simulate step size varied according to the different models we propose in Table1.

Caption for figure 1. Simulation example. The simulation proceeded as follows: an individual bird starts the simulation at time zero and location zero zero. That bird is given five different seeds, with gut retention time (GRT_I), where GRT is gamma distributed. The animal is allowed to move freely across the landscape where at each time point T_i, a movement angle is sampled ~uniform(360) (think of greek letter) and a step size is randomly selected d_i, where D ~ f(x), where f(x) is the probability density function of one of the models we've selected. The movement process is repeated and at T_i == GRT_i a seed is dropped at the bird's location (x_i, y_i) and therefore that seed's final locations is (x_i, y_i). The simualation ends at T_i = max(GRT_1:5), and seed dispersal distances are calculated from the origin (0,0) to each seed's location (x_i, y_i).

# Simulations at different organizational levels

Our main hypothesis is about the incorporation of variation in animal movement at different organizational levels, meaning how are the differences in individual animal movement affecting population level estimates of seed dispersal. How does the underlying variation in individual animal movement impact our estimates of long distance seed dispersal. We explored the effects of individual differences in movement by modifying our base simulation, described above, in which the probability density function used to sample step sizes was different between population, individual, or family level simulations. What we consider as population level simulations are those where all birds share the same probability density to sample step sizes from, while for individual and family level simulations we use the same overall pdf, but allow for variation in the parameters specified for such function, where each set of parameters corresponds to a specific individual or family group. We also performed simulations in which we allowed for variation not only of parameters, but also of the overall structure of the pdf for sampling step sizes, this means that each individual or family group was allowed to have it's own pdf and its own set of parameters (Box 1 or table 1 with description of models, either the movement or simulation). We used four different probability density functions for sampling the animal movement step size, these included an exponential distribution [@jones_closing_2017], gamma distribution [citation], Weibull distribution[citation], and lognormal distribution [@levey_modelling_2008]. These four distributions are leptokurtic, and they vary in the fatness of their tails, thus allowing for different levels of variation in step size. **should add number of runs? or mention that that is in appendix B**

# Parameterization

We focused our study on the many-banded araçari (*Pterglossus pluricinctus*), a small toucan, and its role as a frugivore of the *Virola flexuosa* tree. A subset of animal movement data collected in a previous study [@holbrook_home_2011] was selected and used towards our models and simulations. In this previous study, araçari home ranges, average movement distances, patterns and movement rates were estimated for tracked birds over a period of four years, from 2001 to 2005, in the Ecuadorian Amazon rainforest. Further details on field methods can be found in Holbrook [-@holbrook_home_2011]. Although this previous study included two larger Ramphastids as well, we used data only from the many-banded araçari, because a higher number of recorded locations were available. In addition to this, the home range of this species was found to be the largest for all the species studied, and therefore it was better suited to study long-distance seed dispersal events.

### Animal movement probability distributions  
What data am I using for this parameterization? It is radiotracking data, and the information on continuous trajectories for each tracked bird, to scale those distances into number of meters moved per minute for each tracking segment. I consider a tracking segment as the distance moved between two continuous radio tracking detections. (Check Kimberly's paper), because here we are taking advantage of all the tracking data since there are different tracking interval times. Then, we used that data to fit the four models described above, and did it at the different organizational levels. We considered population level data, as all the data pooled from all individuals tracked, and fit the four distributions to that data to estimate parameters. At the individual level, we fit the four pdfs for the data in respect to each individual. For the family level data, we used only a subset of the total data set to include only individuals for which family level information existed. Because these are social individuals, they tend to move with their group and so we their movements may be correlated. We wanted to explore if there were any differences at the family group levels. We fit the four pdfs to the subset of this data, as the subset population level, and then fit each distribution to the data associated to each of the family groups. We used the parameters estimated from these fits to sample step sizes for our simulation models. We also evaluated the fit of each of these distributions visually through qq plots (found in appendix A), and further details on data organization and model fitting can be found in the online repository.

### Model selection? for the step size

$$AIC{_{indlevel}} = 2 \cdot \sum \left ( number parameters \right ) + 2 \cdot \sum \left ( negative log likelihood \right )$$

$$ AICc{_{indlevel}} =  AIC{_{indlevel}} + (2 \cdot \sum{nparams}  \cdot \left(\sum{nparams}+1\right)/\left(\sum{observations} - \left(\sum{nparams} - 1\right)\right)$$

$$BIC_{indlevel} = \ln{\sum observations} \cdot \sum{number parameters} + 2\cdot \sum {negative log likelihood}$$

# Statistics associated to the simulations
We analyzed simulation outputs and had seed dispersal distance as our main focus. We calculated summary statistics for each of the dispersal distances, and estimated the percentage of long distance dispersal events. Using the 500 meter threshold which was determined in previous work.

## Results
### Animal movement rate models

We compared the fit of the different probability distributions to animal movement rate data with Akaike's Information Criteria(AIC) and it's corrected version for small sample sizes (AICc), we also used the Bayesian Information Criteria (BIC) to compare these model fittings. At the population level, all information criteria concur and show that the lognormal model fits the data for movement rates the best, with the weibull model next, see table 1. However, as we visually assessed fit of these models with QQ plots and goodness-of-fit statistics (see the supplemet), the lognormal distribution model greatly overestimates the data towards the tail, whereas all the other models are significantly more conservative and tend to underestimate the data.

At the individual level, we observed similar outcomes, where the lognormal distribution consistently overestimated the data, whereas the other distributions considered did not. When comparing across distribution models at the individual level, AIC and AICc showed the lognormal fit as the best one followed by the exponential model, whereas the BIC had the exponential as the best model followed by the lognormal.

When comparing the population level models with the ones at the individual level, we found that information criteria differed. In the case of AIC and AICc, these consistently categorized the individual level models as best fitting when compared to their equivalent distribution model at the population level. However, BIC values show that only when we consider the exponential distribution model does the individual level do a better fitting than the population, with all other models having higher BIC values for the models that include individual heterogeneity.

The mixed/heterogeneous model considered the best fitting distribution for each individual, and then selected given distribution to be included in the general model that includes movement for all the individuals and calculates the information criteria value. Using different information criteria yielded different best fitting models (see supplement) for each individual, which changed the distributions included on the overall model, therefore producing two different mixed/heterogeneous modelsd with individual variation, one associated to the AIC/AICc target distributions and a different one for BIC. However, all information criteria agreed on nine out of twelve individual models. In the case of AIC and AICc four individuals were better described by an exponential distribution, three by a Weibull distribution and five by a lognormal distribution. When using BIC as the model selection tool, seven individuals were described by an exponential distribution, four by a lognormal and one by the Weibull distribution. The overall AIC, AICc and BIC values were calculated for each of the mixed/heterogenous model and compared to the previous fits, where the first five results are shown in the following table (the rest can be seen in the supplement). This shows that only the lognormal model at the population level is better than the mixed/heterogeneous model by <2 BIC units.

### Seed Dispersal Distances

We analyzed the seed dispersal distances produced by the simulation runs and there were clear differences between the simulation models. These differences clear between models that used different distributions for movement rates, whether or not they consider individual variation, the seed dispersal distances under the lognormal distribution were extremely high with several outliers for dispersal distance. The exponential, Gamma and Weibull models for movement rates produced similar seed dispersal distance distributions at the population level. All the simulations showed that models that included individual variation in animal movement rates had a higher number of long distance dispersal events.

## Discussion

added Oct 2020: call for the need for more complex animal movement models when running seed dispersal simulations. Even the posibility of mixed distribution models, where not only we allow for each individual or fmailiy group to have it's own model, but also use multiple distributions for each. The short movements performed by these animals may be better described by a different distribution than those performed at longer intervals.

Considering animal movement in this simple framework of movement rates allows us to easily compare between individuals and how including intraspecific variation can change our estimates and fits of different models. It is known that because of this individual variation, no one distribution will fit all individuals, and therefore approaches with mixed distributions or that select different distributions for each individual are important if we want to understand the relationship between frugivore movement and seed dispersal [@russo_incorporating_2006]. Previous research has focused on exploring the effects of lanscape fragmentation, heterogeneity [@jones_closing_2017, @levey_modelling_2008] or plant aggregation [@pegman_exploring_2017]. However, focus on intraspecific variation in frugivores and their effect on seed dispersal has only recently been studied more closely[@snell_consequences_2019]. The models we have used in this study show that individual variation between individual movement rates exist. There is clear evidence for heterogeneity across individuals with respect to which distribution explains the data best with both, the AIC and the BIC, regardless of whether these information criteria match. We need to move towards developing better animal movement models that incorporate this type of heterogeneity, where we don't just pool tracking data for all individuals, but where we consider the tail end distributions of their movement.

We have also shown that including individual variation in type of distribution for movement rates produces one of the best fitting models to the data. Although in our direct comparisson between mixed models and single distribution models, the lognormal distribution at the population level was the best fitting one, the difference of 1.900455 BIC points is too small for us to reject our mixed model that states heterogeneity across individuals. This small difference points at perhaps how small sample sizes per individual can have an effect on the fits.Using BIC, a difference of approximately 4 BIC points is equivalent to stating significant difference at an 0.05 alpha level, therefore the difference of <2 BIC points is not sufficient for us to reject our mixed models with individual variation. All other models that follow are over 10 BIC units from out mixed/heterogeneous model.

Even though LDD events are rare or infrequent, they have a disproportionately large effect on gene flow and the genetic pool of populations (Jordano 2017). In the case of well mixed populations, long-distance dispersal can cause random genetic loss through drift or have the opposite effect and maintain high genetic variance in populations that initially drifted (Bohrer, Nathan, and Volis 2005). This is important as in the long-term, these genetic consequences of dispersal can have significant effects for the survival of populations. From a metapopulation perspective, LDD events can have significant consequences for species persistence by enhancing genetic variability and as mechanism for survival in spatially and temporally heterogenous environments. In particular, long-distance dispersal events have a higher probability of reaching isolated populations and therefore establishing a connection and maintaining unrelated populations. It is of special interest to focus in highly heterogenous habitats with rapid change in spatial structure where local extinctions are high, since long-distance dispersal can allow persistence of a metapopulation with immigration and emigration based on long distance dispersal. From a more evolutionary perspective, long-distance dispersal can allow for a species to colonize distant habitats and expand its range, which can also lead to differentiation and speciation (Ronce 2007). We have shown that individual differences in frugivore movement can have significant consequences on the number of long-distance seed dispersal events and it remains to be explored how these individual differences can influence population dynamics in the long term and at larger spatial scales.

>>>>

# Methods
## Simulation description

To study the effects of individual variation in animal movement over final dispersal distances of foraged seeds, we developed a spatially-explicit individual-based model in a homogenous landscape consisting of a single source tree. We incorporated animal movement focusing on two characteristics, a movement distance (MD) sampled from a probability density distribution, and a movement angle (MA). In addition to this, we included gut retention time (GRT), the time that ingested seeds stay within the frugivore until they are dropped. The animal was allowed to move freely within the landscape, and GRT determined when the animal would drop a seed. Once a seed was dropped, its location was recorded and seed dispersal distance was estimated as the distance from the origin to the seed's location. The average seed dispersal distance for each simulation run was calculated and used to estimate seed dispersion, calculated as the mean distance of each seed to the average location of all seeds in the run [@jones_closing_2017]. For each simulation run, the simulation started at the source tree, an animal received a specific number of seeds and each seed was assigned a specific GRT sampled from a gamma distribution [@morales_effects_2006]. The simulation was based on one minute time steps, where at each time step a movement distance (MD) and movement angle (MA) would determine the path for the animal to follow. For each seed, the location where it would get dropped was based on the animal's location at the time it reached the GRT for each of the seeds. The total simulation time for each run was determined by the largest GRT sampled for that specific simulation run. Each simulation run focused on one individual animal dispersing seeds from one focus tree. Once all seeds in that run were dropped by the animal, the simulation run ended.

## Study design

To understand the consequences of individual variation in animal movement and how these influence seed dispersal distances, we simulated two different scenarios by varying the movement of individual animals in each of those scenarios. We used our first scenario as our null model, in which all the individuals from the simulation had the same average movement rate ($\lambda_1 = \lambda_2 = ... = \lambda_n$), used as the parameter in the probability density function, and thus their movement distances per unit of time were sampled from the same probability distribution ($f(x;\lambda) = \lambda e^{-\lambda x}$). Our second scenario included individual variation in animal movement by incorporating different movement rates for each individual ($\lambda_1 = \lambda_2 = ... = \lambda_n$), which meant that movement distances would be sampled from probability distributions with different parameters ($f(x;\lambda_i) = \lambda_i e^{-\lambda_i x}$). Considering that not all individuals are independent of each other, and that by belonging to social groups their movement patterns might be correlated, we also ran simulations at the scale of family group variation in animal movement by assigning a different movement rate to each social group in the simulations.

@@

## Parameterization

Where is the data coming from?
What is a movement rate and how are we calculating it? Why are we using that (uneven time intervals)?
Why are we using an exponential distribution and what type of movement is this? It is an uncorrelated random walk, but the step size is different.

We give randomly between 3-5 seeds, because that's what's observed in the field. We are simulating here the trajectory and fate of seeds after only one event of frugivory to make things simpler.

We have a maximum landscape area, based on the home range size of the individuals, we chose a grid, and it's modled as a torus so that there is no edge effect.? But do I want this if we want to explore long distance?

Gut retention time based on data and using a gamma distribution from morales paper.

## Seed dispersal distance and agregation metrics

How we calculated these distances, the averages and SD, and then how we calculated the dispersion, long distance dispersal events.

## Seed dispersal kernel fitting

Look into this, but probably use a Weibull distribution like morales and carlo, and then compare the parameters to describe them. An alternative approach would be to use an extreme distribution like the garcia paper to describe this.

# Results

## Seed dispersal distances and long distance dispersal

## Seed dispersion aggregation

## Extreme distributions to fit dispersal

# Discussion
